\thetitle

### 2020-11-09 sun_phases/spin_qudit_tomo.tex
added comment about state-dependent error bound

[\red{todo: add a state-dependent error bound}]

@@

Rather than the statistical errors $\epsilon_{i\mu} \equiv \tilde\Pi_{i\mu} - \bk{\Pi_{i\mu}}_\rho$ in the estimates $\tilde\Pi_{i\mu}$ of the projectors $\Pi_{i\mu}$, we now consider the statistical errors $\epsilon_{i\ell} \equiv \tilde T_{i\ell} - \bk{T_{i\ell}}_\rho$ in the estimates $\tilde T_{i\ell}$ of the transition operators $T_{i\ell} \equiv T_{\v v_i\ell,0}$.
We can expand the transition operators $T_{i\ell}$ as a sum over projectors $\Pi_{i\mu}$ as
\begin{align}
  T_{i\ell} = \sum_\mu t_{\ell\mu} \Pi_{i\mu},
  &&
  t_{\ell\mu} \equiv \bk{\mu|T_{\ell,0}|\mu}
  = \sqrt{\f{2\ell+1}{d}} \bk{s\mu;\ell,0|s\mu},
\end{align}
and likewise $\tilde T_{i\ell} = \sum_\mu t_{\ell\mu} \tilde\Pi_{i\mu}$.
The covariance between errors in the transition operator basis is then
\begin{align}
  \bbk{\epsilon_{i\ell} \epsilon_{i'\ell'}}
  = \sum_{\mu,\mu'} t_{\ell\mu} t_{\ell'\mu'}
  \bbk{\epsilon_{i\mu} \epsilon_{i'\mu'}}
  = \delta_{ii'} \times \f1n \sum_{\mu,\mu'} t_{\ell\mu} t_{\ell'\mu'}
  \p{\delta_{\mu\mu'} p_\mu^i - p_\mu^i p_{\mu'}^i}.
\end{align}
In order to establish an upper bound on reconstruction error for an unknown quantum state, we will need an upper bound on the variance $\bbk{\epsilon_{i\ell}^2}$ in particular.
To this end, we observe that
\begin{align}
  \bbk{\epsilon_{i\ell}^2} = \f1n \times \sigma_{p^i}^2\p{t_\ell},
\end{align}
where $t_\ell \equiv \p{t_{\ell,-s},t_{\ell,-s+1},\cdots,t_{\ell s}}$ is a vector of all $t_{\ell\mu}$, and
\begin{align}
  \sigma_p^2\p{X}
  \equiv \sum_\mu p_\mu X_\mu^2 - \p{\sum_\mu p_\mu X_\mu}^2
  = \sum_\mu p_\mu \p{X_\mu - \sum_\nu p_\nu X_\nu}^2
\end{align}
is the variance of $X$ when sampled according to the probability distribution $p$.
The variance $\sigma_p^2\p{X}$ is maximal when $p$ has equal weight on the largest and smallest values of $X$, which implies that
\begin{align}
  \sigma_p^2\p{t_\ell} \le \Gamma_\ell^2,
  &&
  \Gamma_\ell \equiv \f{\max_\mu t_{\ell\mu} - \min_\mu t_{\ell\mu}}{2}.
\end{align}
Altogether, using additionally the fact that $\sigma_p^2\p{t_0}=0$ (because all $t_{0,\mu}=1/\sqrt{d}$) we can bound
\begin{align}
  \bbk{\epsilon_{i\ell}^2}
  \le \f1n \times \Gamma_\ell^2 \times \p{1 - \delta_{\ell,0}}
  = \f{2\ell+1}{nd} \times \gamma_\ell^2
  \times \p{1 - \delta_{\ell,0}},
  \label{eq:variance_bound}
\end{align}
where $\gamma_\ell$ is defined using Clebsch-Gordan coefficients:
\begin{align}
  \gamma_\ell \equiv \f{\max_\mu\tau_{\ell\mu} - \min_\mu\tau_{\ell\mu}}{2},
  &&
  \tau_{\ell\mu} \equiv \bk{s\mu;\ell,0|s\mu}.
\end{align}
Note that the bound in Eq.~\eqref{eq:variance_bound} is tight, as equality is achieved by the mixed state
\begin{align}
  \rho^\star
  \equiv \f12 \p{\Pi_{\v v_i,\mu_{\t{max}}} + \Pi_{\v v_i,\mu_{\t{min}}}},
\end{align}
where $\mu_{\t{max}}$ ($\mu_{\t{min}}$) is any index that maximizes (minimizes) $\tau_{\ell\mu}$.

@@

>>>>

As a final comment, we note that the statistical error bound in Eq.~\eqref{eq:bound} is a ``worst case'' bound that is agnostic to the state $\rho$.
The actual statistical error for any particular state may be considerably smaller.
Written out in full (and derived in Appendix \ref{sec:error}),
\begin{align}
  \Bbk{\norm{\tilde\rho_V-\rho}^2}
  = \f1n \sum_{i,\ell} \sigma_\rho^2\p{T_{\v v_i\ell,0}}
  \sum_m \abs{q_{\ell mi}^V}^2,
  \label{eq:error}
\end{align}
where $i$ indexes an axis $\v v_i\in V$; $\sigma_\rho^2\p{X} \equiv \bk{X^2}_\rho - \bk{X}_\rho^2$ is the variance of $X$; and $\v q_{\ell m}^V$ is a right singular vector of $M_{V\ell}$ corresponding to the singular value $M^V_{\ell m}$, with normalization $\abs{\v q_{\ell m}^V}=1/M^V_{\ell m}$.
While the error in Eq.~\eqref{eq:error} cannot be known exactly without knowing $\rho$, this error can be estimated by computing the right-hand side of Eq.~\eqref{eq:error} with the replacement $\rho\to\tilde\rho_V$.

@@

Rather than the statistical errors $\epsilon_{i\mu} \equiv \tilde\Pi_{i\mu} - \bk{\Pi_{i\mu}}_\rho$ in the estimates $\tilde\Pi_{i\mu}$ of the projectors $\Pi_{i\mu}$, we now consider the statistical errors $\epsilon_{i\ell} \equiv \tilde T_{i\ell} - \bk{T_{i\ell}}_\rho$ in the estimates $\tilde T_{i\ell}$ of the transition operators $T_{i\ell} \equiv T_{\v v_i\ell,0}$.
We can expand the transition operators $T_{i\ell}$ as a sum over projectors $\Pi_{i\mu}$ as
\begin{align}
  T_{i\ell} = \sum_\mu t_{\ell\mu} \Pi_{i\mu},
  &&
  t_{\ell\mu} \equiv \bk{\mu|T_{\ell,0}|\mu}
  = \sqrt{\f{2\ell+1}{d}} \bk{s\mu;\ell,0|s\mu},
\end{align}
and likewise $\tilde T_{i\ell} = \sum_\mu t_{\ell\mu} \tilde\Pi_{i\mu}$.
The covariance between errors in the transition operator basis is then
\begin{align}
  \bbk{\epsilon_{i\ell} \epsilon_{i'\ell'}}
  = \sum_{\mu,\mu'} t_{\ell\mu} t_{\ell'\mu'}
  \bbk{\epsilon_{i\mu} \epsilon_{i'\mu'}}
  = \delta_{ii'} \times \f1n \sum_{\mu,\mu'} t_{\ell\mu} t_{\ell'\mu'}
  \p{\delta_{\mu\mu'} p_\mu^i - p_\mu^i p_{\mu'}^i}
  = \delta_{ii'} \times \f1n \sp{\bk{T_{i\ell} T_{i\ell'}}_\rho
    - \bk{T_{i\ell}}_\rho \bk{T_{i\ell'}}_\rho}.
\end{align}
In order to establish an upper bound on reconstruction error for an unknown quantum state, we will need an upper bound on the variance $\bbk{\epsilon_{i\ell}^2}$ in particular.
To this end, we observe that
\begin{align}
  \bbk{\epsilon_{i\ell}^2} = \f1n \times \sigma_{p^i}^2\p{t_\ell},
\end{align}
where $t_\ell \equiv \p{t_{\ell,-s},t_{\ell,-s+1},\cdots,t_{\ell s}}$ is a vector of all $t_{\ell\mu}$, and
\begin{align}
  \sigma_p^2\p{X}
  \equiv \sum_\mu p_\mu X_\mu^2 - \p{\sum_\mu p_\mu X_\mu}^2
  = \sum_\mu p_\mu \p{X_\mu - \sum_\nu p_\nu X_\nu}^2
\end{align}
is the variance of $X$ when sampled according to the probability distribution $p$.
The variance $\sigma_p^2\p{X}$ is maximal when $p$ has equal weight on the largest and smallest values of $X$, which implies that
\begin{align}
  \sigma_p^2\p{t_\ell} \le \Gamma_\ell^2,
  &&
  \Gamma_\ell \equiv \f{\max_\mu t_{\ell\mu} - \min_\mu t_{\ell\mu}}{2}.
\end{align}
Altogether, using additionally the fact that $\sigma_p^2\p{t_0}=0$ (because all $t_{0,\mu}=1/\sqrt{d}$) we can bound
\begin{align}
  \bbk{\epsilon_{i\ell}^2}
  \le \f1n \times \Gamma_\ell^2 \times \p{1 - \delta_{\ell,0}}
  = \f{2\ell+1}{nd} \times \gamma_\ell^2
  \times \p{1 - \delta_{\ell,0}},
  \label{eq:variance_bound}
\end{align}
where $\gamma_\ell$ is defined using Clebsch-Gordan coefficients:
\begin{align}
  \gamma_\ell \equiv \f{\max_\mu\tau_{\ell\mu} - \min_\mu\tau_{\ell\mu}}{2},
  &&
  \tau_{\ell\mu} \equiv \bk{s\mu;\ell,0|s\mu}.
\end{align}
Note that the bound in Eq.~\eqref{eq:variance_bound} is tight, as equality is achieved by the mixed state
\begin{align}
  \rho^\star
  \equiv \f12 \p{\Pi_{\v v_i,\mu_{\t{max}}} + \Pi_{\v v_i,\mu_{\t{min}}}},
\end{align}
where $\mu_{\t{max}}$ ($\mu_{\t{min}}$) is any index that maximizes (minimizes) $\tau_{\ell\mu}$.

@@

As a final comment, we note that the statistical error bound in \eqref{eq:bound_full} is a ``worst-case'' bound that is agnostic to the state $\rho$.
The actual statistical error for any particular state may be considerably smaller, and can be obtained by substituting the exact expression for $\bbk{\epsilon_{i\ell}^2}$:
\begin{align}
  \Bbk{\norm{\tilde\rho_V-\rho}^2}
  = \sum_{i,\ell,m} \abs{q^V_{\ell mi}}^2 \bbk{\epsilon_{i\ell}^2}
  = \f1n \sum_{i,\ell} \sigma_\rho^2\p{T_{i\ell}}
  \sum_m \abs{q^V_{\ell mi}}^2,
  \label{eq:error_full}
\end{align}
where $\sigma_\rho^2\p{X}\equiv\bk{X^2}_\rho-\bk{X}_\rho^2$ is the variance of $X$.
While the error in Eq.~\eqref{eq:error_full} cannot be known exactly without knowing $\rho$, this error can be estimated by computing the right-hand side of Eq.~\eqref{eq:error_full} with the replacement $\rho\to\tilde\rho_V$.
